---
title: "Cleaning OpenPose gait keypoints with PoseFixeR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cleaning OpenPose gait keypoints with PoseFixeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r setup}
library(PoseFixeR)
```

`PoseFixeR` cleans two-dimensional, video-based gait recordings produced by
[OpenPose](https://github.com/CMU-Perceptual-Computing-Lab/openpose). In an
uncontrolled filming environment the pose estimator makes mistakes --- it swaps
the left and right leg, produces anatomically impossible segment lengths or
joint angles, jumps between frames, or reports low-confidence detections.
`posefixer()` detects those anomalies, removes the offending keypoints, and then
imputes and smooths the trajectories. This vignette walks through the whole
workflow on a small **synthetic** recording so it runs offline in a few seconds;
with real data you would simply point `import_json()` at your OpenPose output.

## A synthetic recording

OpenPose writes one JSON file per video frame. The reader in this package
expects each file to hold a flat vector of `18 * 3` numbers --- an `X`, `Y` and
confidence value for each of the 18 body keypoints. We fabricate 40 frames of a
side-on walker whose legs swing in anti-phase, and write them to a temporary
directory to mimic a real OpenPose run.

```{r synthesize}
parts <- c("nose","neck","Rshoulder","Relbow","Rwrist","Lshoulder","Lelbow",
           "Lwrist","Rhip","Rknee","Rankle","Lhip","Lknee","Lankle",
           "Reye","Leye","Rear","Lear")
bx <- c(0,0,-.25,-.3,-.32,.25,.3,.32,-.18,-.2,-.22,.18,.2,.22,-.05,.05,-.1,.1)
by <- c(-.3,0,.05,.5,.95,.05,.5,.95,1,1.65,2.28,1,1.65,2.28,-.35,-.35,-.3,-.3)

n <- 40
swing <- 0.45 * sin(seq(0, 4 * pi, length.out = n))
dir <- file.path(tempdir(), "openpose_json")
dir.create(dir, showWarnings = FALSE)

files <- character(n)
for (i in seq_len(n)) {
  fx <- bx; fy <- by
  fx[11] <- bx[11] - swing[i];       fx[14] <- bx[14] + swing[i]        # ankles
  fx[10] <- bx[10] - 0.5 * swing[i]; fx[13] <- bx[13] + 0.5 * swing[i]  # knees
  frame <- as.numeric(t(cbind(fx, fy, 1)))     # row-major X, Y, P per keypoint
  files[i] <- file.path(dir, sprintf("frame_%04d.json", i))
  writeLines(rjson::toJSON(frame), files[i])
}
```

## Import and reshape

`import_json()` reads a batch of frame files into a list of
keypoint-by-`c("X", "Y", "P")` matrices; `reshaping()` stacks them into separate
`x`, `y` and confidence matrices with one row per frame and one column per
keypoint.

```{r import}
keypoints <- import_json(sort(files))
length(keypoints)        # one matrix per frame
dim(keypoints[[1]])      # 18 keypoints x (X, Y, P)

raw <- reshaping(keypoints)
dim(raw$x)               # frames x keypoints
```

## Normalise

`normalize()` re-expresses every keypoint relative to the neck and divides by the
hip-neck distance, so that camera position and distance to the subject drop out.

```{r normalize}
pose <- normalize(raw)
range(pose$x[, "neck"])  # the neck sits at the origin
```

## Inspect segment lengths and joint angles

A few helpers read directly off the reshaped data. `distance()` returns the
Euclidean distance between two keypoints (a segment length), while `hdegree()`
and `kdegree()` return hip and knee joint angles from three keypoints.

```{r features}
thigh <- distance(pose, "Rhip", "Rknee", 1:n)   # right thigh length per frame
summary(thigh)

kdegree(pose, 9, 10, 11, 10)   # right knee angle in frame 10
hdegree(pose, 2, 9, 10, 10)    # right hip angle in frame 10
```

`anomaly()` flags values more than `k` standard deviations from the mean of a
numeric series --- handy for spotting an implausible segment length or stride
width before correction.

```{r anomaly}
gap <- distance(pose, "Rankle", "Lankle", 1:n)
flags <- anomaly(gap, k = 2)
sum(flags$caselist)            # number of flagged frames
```

## Detect and correct anomalies

`posefixer()` is the main entry point. It runs the full
detect-correct-impute-smooth workflow and returns the cleaned trajectories in
the same `x`/`y`/`p` structure. The `race` argument selects the reference
anatomical proportions and `fps` is the frame rate of the source video. If a
recording fails the quality criteria the function returns `NA` instead.

```{r posefixer}
fixed <- posefixer(pose, race = 1, fps = 25)
str(fixed, max.level = 1)
```

## Visualise

`plt.segment()` draws the skeleton for one frame on top of a scatter of its
keypoints, and `diplay_frames()` lays out a sequence of frames.

```{r plot-frame}
plot(fixed$x[10, ], fixed$y[10, ], ylim = c(2.5, -0.5), asp = 1, pch = 19,
     xlab = "x", ylab = "y", main = "Corrected pose, frame 10")
plt.segment(fixed$x[10, ], fixed$y[10, ])
```

```{r plot-sequence}
op <- graphics::par(no.readonly = TRUE)
diplay_frames(list(x = fixed$x[c(1, 11, 21, 31), ],
                   y = fixed$y[c(1, 11, 21, 31), ]),
              layout = c(1, 4), xlim = c(-1, 1), ylim = c(2.5, -0.5))
graphics::par(op)
```

## References

Sugiyama Y, Uno K, Matsui Y (2023). Types of anomalies in two-dimensional
video-based gait analysis in uncontrolled environments. *PLOS Computational
Biology* 19:e1009989. <https://doi.org/10.1371/journal.pcbi.1009989>
